---
title: "OutSplice"
author: 
- "Joseph Bendik"
- "Sandhya Kalavacherla"
- "Michael Considine"
- "Bahman Afsari"
- "Michael F. Ochs"
- "Joseph Califano"
- "Daria A. Gaykalova"
- "Elana Fertig" 
- "Theresa Guo"
package: "OutSplice"
output: 
    BiocStyle::html_document:
      toc: true
      toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Find Splicing Outliers in Tumor Samples with OutSplice}
  %\VignetteEncoding{UTF-8}  
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

`r Biocpkg("OutSplice")` is a package that identifies aberrant splicing events in tumor samples when compared to normal samples using outlier statistics to better characterize sporadic events occurring in heterogenous cancer samples. Events are classified as skipping, insertion, or deletion events. Additionally, outlier expression of each splicing event is called per sample allowing for evaluation of splicing burden of individual samples. This package is specifically designed to analyze RNA sequencing data provided by the user; additional functions have also been added for direct analysis from the Cancer Genome Atlas (TCGA). Packages in Bioconductor, such as the `r Biocpkg("psichomics")` package, perform similar analyses on TCGA or user-generated data, however they are not specifically designed to identify cancer specific events. Because of the heterogeneity of tumors in the context of cancer biology, outlier statistics rather than mean comparisons are used to identify aberrant events, where outliers are classified on a per sample basis; this allows the algorithm to discern splicing events with outlier expression within that may occur within a subset of tumor samples when compared to normal conditions. Therefore, this package generates a matrix of splicing outliers, which are associated with splice junctions that are either significantly over or under-expressed compared to the distribution of expression in normal tissue. The `r Biocpkg("FRASER")` package also offers functionality to determine splicing outliers, however, `r Biocpkg("OutSplice")` utilizes a rank sum outlier approach instead of a beta binomial model. This allows for users to eliminate less biologically relevant events by setting a minimum level of event expression. Additionally, `r Biocpkg("OutSplice")` has an easy to use work-flow that makes it accessible for researchers of varying experience with bioinformatics tools. With `r Biocpkg("OutSplice")`, the junction normalization, filtering, outlier calling, and statistical analysis are done in one function call, simplifying this process for the user. `r Biocpkg("OutSplice")` is also capable of automatically calculating the splicing burden in each of the user’s samples and will provide waterfall plots of expression to help visualize individual splicing events. Overall, this package is novel in that it determines differential splicing burdens between tumors and normal samples and characterizes the nature of splicing outliers in a quick and user friendly fashion.

### Functionality

The main functions of `r Biocpkg("OutSplice")` achieve the following for either user 
provided data or data provided from the TCGA.

1. Junction normalization
2. Outlier analysis
3. Determination of a junctional outlier as a skipping, insertion, or deletion
4. Calculation of splicing burden
5. Plotting expression levels

### Minimum required packages

`r Biocpkg("OutSplice")` will import the below packages and their dependencies so please ensure they are installed 
before using the software:

1. AnnotationDbi
2. GenomicRanges
3. GenomicFeatures
4. IRanges
5. org.Hs.eg.db
6. Repitools
7. TxDb.Hsapiens.UCSC.hg19.knownGene
8. TxDb.Hsapiens.UCSC.hg38.knownGene
9. S4Vectors

### Installation from Bioconductor
The `r Biocpkg("OutSplice")` package is available at https://bioconductor.org and can be 
installed via BiocManager::install:

```{r, echo = FALSE, eval=FALSE}
if (!require("BiocManager")) {
    install.packages("BiocManager")
}
BiocManager::install("OutSplice")
```

A package only needs to be installed once. Load the package into an R session with:

```{r, message=FALSE}
library(OutSplice)
```

# Inputs

### outspliceAnalysis/outspliceTCGA

Full Sample data for `r Biocpkg("OutSplice")` can be downloaded from The Cancer Genome Atlas (TCGA).

The outspliceAnalysis function has 6 required inputs and 6 optional Inputs.
The outspliceTCGA function has 5 required inputs and 6 optional Inputs and 
should only be used if the data is in TCGA format.

In the below examples we run outspliceAnalysis and outspliceTCGA on a subset of Head 
and Neck squamous cell carcinoma data obtained from The Broad Institute TCGA 
GDAC Firehose database.

```{r, results="hide", warning=FALSE, messages = FALSE}
junction <- system.file("extdata", "HNSC_junctions.txt.gz", package = "OutSplice")
gene_expr <- system.file("extdata", "HNSC_genes_normalized.txt.gz", package = "OutSplice")
rawcounts <- system.file("extdata", "Total_Rawcounts.txt", package = "OutSplice")
sample_labels <- system.file("extdata", "HNSC_pheno_table.txt", package = "OutSplice")
output_file_prefix <- "OutSplice_Example"
TxDb_hg19 <- "TxDb.Hsapiens.UCSC.hg19.knownGene"
dir <- paste0(tempdir(), "/")
message("Output is located at: ", dir)

outspliceAnalysis(junction, gene_expr, rawcounts, sample_labels, output_file_prefix, dir, filterSex = TRUE, annotation = "org.Hs.eg.db", TxDb = TxDb_hg19, offsets_value = 0.00001, correction_setting = "fdr", p_value = 0.05)
```

```{r, results="hide", warning=FALSE, messages = FALSE}
junction <- system.file("extdata", "TCGA_HNSC_junctions.txt.gz", package = "OutSplice")
gene_expr <- system.file("extdata", "TCGA_HNSC_genes_normalized.txt.gz", package = "OutSplice")
rawcounts <- system.file("extdata", "Total_Rawcounts.txt", package = "OutSplice")
output_file_prefix <- "TCGA_OutSplice_Example"
dir <- paste0(tempdir(), "/")
message("Output is located at: ", dir)

outspliceTCGA(junction, gene_expr, rawcounts, output_file_prefix, dir, filterSex = TRUE, annotation = "org.Hs.eg.db", TxDb = "TxDb.Hsapiens.UCSC.hg19.knownGene", offsets_value = 0.00001, correction_setting = "fdr", p_value = 0.05)
```
  
  -junction: filepath to a matrix of junction raw read counts
  
  -gene_expr: filepath to a matrix of normalized gene expression data. RSEM quartile normalized data is recommended.
  
  -rawcounts: filepath to a matrix of raw read counts for each gene. Can either be per gene, or a summed total for each sample.
  
  -sample_labels: filepath to a matrix of phenotypes (in Section 2.0.2)
  
  -output_file_prefix: user defined string for what the prefix of the output file should be named
  
  -dir: string containing the desired output path. Ex) "~/my_outsplice_output/"
  
  -filtersex: ignores sex chromosomes when generating results. True by default [OPTIONAL]
  
  -annotation: the bioconductor package containing the annotations to use. Uses the human genome by default (in Section 3.0.4) [OPTIONAL]
  
  -TxDb: the bioconductor package containing the transcript annotations to use. The outspliceAnalysis function uses the hg38 genome by default, and outspliceTCGA uses hg19 by default (in Section 3.0.4). [OPTIONAL]
  
  -offsets_value: the normalized junction expression threshold. Uses 0.00001 by default. (in Section 3.0.6) [OPTIONAL]
  
  -correction_setting: the correction value to be used for p-value adjustment during Fisher analyses. The available options are: "holm", "hochberg", "hommel", "bonferroni", "BH", "BY", "fdr", and "none". Uses 'fdr' by default. [OPTIONAL]
  
  -p_value: the significance threshold to use during Fisher analyses. Uses 0.05 by default. [OPTIONAL]

This algorithm is compatible with any organism provided the genome object, 
genome wide annotation, and transcript annotations exist as packages on Bioconductor.

### Phenotype matrix 

If using the outspliceAnalysis function, users must provide a phenotype matrix, 
designating which samples in the junction file belong to the tumor group 
(labeled as "T") and the normal group (labeled as "F"). Please ensure the 
matrix file contains a header row with the first column designating the sample 
names, and the second column designating the phenotype. If using TCGA data, 
the two phenotypes are "tumor" and "normal." outspliceTCGA can automatically 
infer the phenotype of TCGA data using the sample names. Only if the phenotype 
matrix has more than 10 control samples, the program proceeds. 

### plotJunctionData

After running outspliceAnalysis or outspliceTCGA, the user can create plots of specified
junctions using the plotJunctionData function.

The function plotJunctionData has 1 required input and 10 optional inputs

In the below example we can plot the expression levels for a "skipping" junction
event on the ECM1 gene based on our example data from the TCGA.

```{r, results="hide", warning=FALSE, messages=FALSE}
data_file <- system.file("extdata", "OutSplice_Example_2023-01-06.RDa", package = "OutSplice")
ecm1_junc <- "chr1:150483674-150483933"
pdf <- "ecm1_expression.pdf"
pdf_output <- paste0(tempdir(), "/", pdf)
message("Output is located at: ", pdf_output)

plotJunctionData(data_file, NUMBER = 1, junctions = ecm1_junc, tail = NULL, p_value = 0.05, GENE = FALSE, SYMBOL = NULL, makepdf = TRUE, pdffile = pdf_output, tumcol = "red", normcol = "blue")
```

  -data_file: filepath to an R Data file containing output from the outspliceAnalysis or outspliceTCGA functions.

  -NUMBER: number of junctions to plot.  This can be top number of junctions (over or under expressed), or can be specific junctions in a list.  Default is 1 [OPTIONAL]

 -junctions: you can input the specific junction you want to graph (or vector of junctions). Default is NULL [OPTIONAL]

 -tail: you can specify if you want top over or under expressed with tail='RIGHT' for junctions overexpressed in tumors, or tail='LEFT' for junctions underexpressed in tumors.  Default is NULL [OPTIONAL]

 -p_value: p-value threshold to use when plotting the top over or under-expressed junctions with "tail". Default is 0.05 [OPTIONAL]

 -GENE: Pick junctions based on a specific gene. TRUE means you will pick all the junctions mapping to a certain gene.  FALSE means you do not pick based on the gene. Default is False. [OPTIONAL]

 -SYMBOL: HGNC SYMBOL of gene you want to graph [OPTIONAL]

 -makepdf: Save graphs to a pdf?  Default is no/FALSE [OPTIONAL]

 -pdffile: if you want to save to a pdf, you will need to write the file path [OPTIONAL when makepdf = F]

 -tumcol: color for tumors on graph. Default is red [OPTIONAL]

 -normcol: color for normals on graph. Default is blue [OPTIONAL]

# Methodology

The below sections describe the processes used in the above functions.

### Junction RPM normalization

The program automatically normalizes the junction counts by dividing the 
junction counts by the total raw counts and then dividing each count by 10^6 
to generate RPM junction data.

### OGSA initial filtering

The dotheogsa function from the Bioconductor package OGSA is sourced to remove 
junctions that may not be biologically relevant due to low expression or that have any difference between tumor and normal. 
In this package, we set a 0.1 RPM expression threshold for pre-filtering. 

### OGSA outlier analysis

The dotheogsa function is again employed to determine splicing events as 
outliers, which are defined as any normalized junctions that are two standard 
deviations above or below the median of the normal distribution. A Fisher exact 
test is used to determine which junctions are significantly over or under 
expressed in tumors compared to the normal samples.

### Genomic references

The Bioconductor GenomicRanges packages are used to assign each junction to a 
known gene. The user has the option in the main function to input which genome 
and its associated Bioconductor packages to use as the reference. These Bioconductor 
packages should include the genome object, annotations, and transcript annotations.

Ex) For mouse genomes aligned to mm39, install and load: 
`r Biocpkg("Mus.musculus")` (genome object), `r Biocpkg("org.Mm.eg.db")` (annotations), 
and `r Biocpkg("TxDb.Mmusculus.UCSC.mm39.refGene")` (transcript annotations) using the library() function. Then, when using the `r Biocpkg("OutSplice")` functions, specify `r Biocpkg("org.Mm.eg.db")` for the annotation argument, 
and `r Biocpkg("TxDb.Mmusculus.UCSC.mm39.refGene")` for the TxDb argument.

This process will also work for any genome provided the correct annotation 
package, its corresponding transcript annotation package, and their dependencies 
from Bioconductor are provided and loaded.

Using this genomic assignment, the dotheogsa function determines insertion, skipping, or deletion events based on the following criteria:

insertion: junction that starts outside a known exon
skipping: junction that skips over a known exon
deletion: junction that is inside a known exon but not as its start or end

### Junction expression normalization

Junction expression is normalized based on its corresponding gene expression 
from the gene_expr input. This is achieved by dividing the junction RPM data by the 
normalized gene expression counts from a junction's corresponding gene. If a junction is aligned
to more than one gene, then the first gene will be the one selected for the normalization.

### Filter by expression via offsets

Offsets, which the user can specify, sets a minimum value relative to the normal
samples in order to call a junction an outlier. The goal is to remove data with 
low expression that may not be biologically relevant. In this example example, an 
outlier junction must have a normalized expression greater than 0.00001 in 
order to be called an outlier. Any outliers with expressions below this value 
are too low to be relevant for the analysis in this example.

### Splice Burden Calculation

Sums the number of splicing events in each sample that were marked as a TRUE 
outlier for both over-expressed and under-expressed events. The total number of
outliers is then calculated as the sum of the over and under-expressed outliers.

### Junction Plotting

Creates bar and waterfall plots of junction expression in both the tumor and
normal samples. The data for these plots comes from the raw junction input, the
gene expression values to reflect overall gene expression, and the junction
expression normalized by gene expression.

# Outputs

### outSpliceAnalysis/outSpliceTCGA

Outputs an RData file with the following data:

  -ASE.type: significant junction events labeled by type (skipping, insertion, or deletion)
  
  -FisherAnalyses: Matrix of junction events containing the number of under/over-expressed outliers in the tumor group (Num_UE_Outliers/Num_OE_Outliers), the Fisher p-value for under/over-expressed events (FisherP1/FisherP2), and a ranking of the under/over expressed events (UE_Rank/OE_Rank)
  
  -geneAnnotations: object containing gene names corresponding to each junction region
  
  -junc.Outliers: list containing the logical matrices TumorOverExpression and TumorUnderExpression. "True" indicates an over-expressed event in TumorOverExpression, or an under-expressed event in TumorUnderExpression.
  
  -junc.RPM: junction counts in reads per million following a division of the junction counts input by the total rawcounts for each sample
  
  -junc.RPM.norm: junction counts normalized by each event's total gene expression value
  
  -gene_expr: gene expression values for each junction event
  
  -splice_burden: matrix containing the number of Fisher-P significant over-expressed, under-expressed, and total number of outliers per sample
  
  -NORM.gene_expr.norm: Median of junction data normalized by gene expression for normal samples only (Used for Junction Plotting Only)
  
  -pheno: Phenotypes of Samples (Tumor or Normal)
  
  -pvalues: Junction Fisher P-values
  
Tab-separated text files are also produced for the following data:

  -ASE.type: event_types.txt
  
  -FisherAnalyses: FisherAnalyses.txt
  
  -geneAnnotations: gene_annotations.txt
  
  -junc.Outliers: TumorOverExpression.txt, TumorUnderExpression.txt
  
  -splice_burden: splice_burden.txt
  
### plotJunctionData

Outputs junction expression plots of user specified junctions as defined in 
Section 3.0.8. Plots can be saved to a user defined pdf file.
  
# References

Cancer Genome Atlas Network. Comprehensive genomic characterization of head and neck squamous cell carcinomas. Nature. 2015 Jan 29;517(7536):576-82. doi: 10.1038/nature14129. PMID: 25631445; PMCID: PMC4311405.

Guo T, Sakai A, Afsari B, Considine M, Danilova L, Favorov AV, Yegnasubramanian S, Kelley DZ, Flam E, Ha PK, Khan Z, Wheelan SJ, Gutkind JS, Fertig EJ, Gaykalova DA, Califano J. A Novel Functional Splice Variant of AKT3 Defined by Analysis of Alternative Splice Expression in HPV-Positive Oropharyngeal Cancers. Cancer Res. 2017 Oct 1;77(19):5248-5258. doi: 10.1158/0008-5472.CAN-16-3106. Epub 2017 Jul 21. PMID: 28733453; PMCID: PMC6042297.

Liu C, Guo T, Sakai A, Ren S, Fukusumi T, Ando M, Sadat S, Saito Y, Califano JA. A novel splice variant of LOXL2 promotes progression of human papillomavirus-negative head and neck squamous cell carcinoma. Cancer. 2020 Feb 15;126(4):737-748. doi: 10.1002/cncr.32610. Epub 2019 Nov 13. PMID: 31721164.

Liu C, Guo T, Xu G, Sakai A, Ren S, Fukusumi T, Ando M, Sadat S, Saito Y, Khan Z, Fisch KM, Califano J. Characterization of Alternative Splicing Events in HPV-Negative Head and Neck Squamous Cell Carcinoma Identifies an Oncogenic DOCK5 Variant. Clin Cancer Res. 2018 Oct 15;24(20):5123-5132. doi: 10.1158/1078-0432.CCR-18-0752. Epub 2018 Jun 26. PMID: 29945995; PMCID: PMC6440699.

M. F. Ochs, J. E. Farrar, M. Considine, Y. Wei, S. Meshinchi, and R. J. 
Arceci. Outlier analysis and top scoring pair for integrated data analysis and 
biomarker discovery. IEEE/ACM Trans Comput Biol Bioinform, 11: 520-32, 2014. PMCID: PMC4156935

# Session Info

```{r}
sessionInfo()
```
